---
title: "Penguin body mass by species and sex (R version)"
author: "Karnbir Nijjar"
date: "2026-09-23"
---

Data: [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/),
Palmer Station Antarctica LTER. Released under CC0.

## On the dataset

This dataset has information on penguins of different species that originate on islands in Antarctica. 

## Intent

We are going to investigate this dataset to find out how body mass differs between males and females in each species and how the species fare against one another in this regard. 

## Summoning the libraries

```{r}
#| label: setup
#| message: false
#| warning: false

library(palmerpenguins)
library(dplyr)
library(ggplot2)
library(knitr)
```

## Glimpsing the dataset

Let's glimpse the dataset we will be operating on:

```{r}
glimpse(penguins)
```

## Manipulating the dataset into a filtered version

We are going to clean up the dataset by filtering out rows that do not have a value for body mass or sex. Then, let's glimpse the new dataset penguins_filtered.

```{r}
#| label: filtered

penguins_filtered <- penguins |>
  filter(!is.na(body_mass_g), !is.na(sex))

glimpse(penguins_filtered)
```

Let's see how different the amount of data is between penguins and our penguins_filtered:

```{r}
nrow(penguins)
nrow(penguins_filtered)
```

Looks like our filtering got rid of 11 rows. 


## Average body mass (g) by species and sex

We are going to find average body mass (rounded to the closest non-decimal number) for males and females of every species in the dataset, formatting that information into a table. 

```{r}
#| label: tbl-body-mass-summary
#| tbl-cap: "Average body mass (g) by species and sex"

avg_body_mass_table <- penguins_filtered |>
  group_by(species, sex) |>
  summarise(
    average_body_mass_in_grams = round(mean(body_mass_g)),
    .groups = "drop"
  )

kable(avg_body_mass_table)
```

## Plotting our penguins 

We are going to make a boxplot of the filtered penguins dataset. The plot will visualize the body mass (in grams) of both males and females in each species.  

```{r}
#| label: fig-penguin-mass
#| fig-cap: "Body mass of penguins by species and sex"

ggplot(penguins_filtered, aes(x = species, y = body_mass_g, fill = sex)) +
  geom_boxplot() +
  labs(x = "Species", y = "Body mass (grams)", fill = "Sex") +
  theme_bw()

```

## Takeaways

It seems that males are generally heavier than their female counterparts in each of the three species in the dataset. This is most prominent in Gentoo penguins.  

Out of the three species, Gentoo penguins are the heaviest. Adelie and Chinstrap penguins do not seem to be too different in their average body masses. 



