---
title: "R and Python working together"
author: "Karnbir Nijjar"
date: "2026-09-23"
---

Data: [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/),
Palmer Station Antarctica LTER. Released under CC0.

## On the dataset

This dataset has information on penguins of three species from three islands in Antarctica. 

## Intent

To replicate the penguins body mass by species and sex analysis, this time using Python to generate the averaged data and R to plot the averaged data. 

## Setup 

```{r}
#| label: setup
#| warning: false
#| message: false
library(reticulate)
library(ggplot2)
use_virtualenv("../../.venv", required = TRUE)
```

## Python element 

```{python}
#| label: python-summary
from palmerpenguins import load_penguins
```

```{python}
#| label: load
penguins = load_penguins()
```

```{python}
#| label: filter
penguins_filtered = penguins.dropna(subset = ['body_mass_g' , 'sex'])
```

```{python}
#| label: tbl-body-mass-summary
#| tbl-cap: "Average body mass (g) by species and sex"

avg_body_mass_table = (
  penguins_filtered
  .groupby(['species', 'sex'])['body_mass_g']
  .mean()
  .round()
  .reset_index(name = "average_body_mass_in_grams")
)

avg_body_mass_table
```

## R element

```{r}
#| label: fig-avg-mass
#| fig-cap: "Average body mass by species and sex using both Python and R"

avg_mass_data <- py$avg_body_mass_table

ggplot(avg_mass_data, aes(x = species, y = average_body_mass_in_grams, fill = sex)) +
  geom_col(position = "dodge") +
  labs(x = "Species", y = "Average body mass (grams)", fill = "Sex") +
  theme_bw()

```

## Takeaways 

Gentoo penguins have the highest average body mass compared to the other two species (Adelie and Chinstrap penguins have very similar average body masses). Across all species, males have larger average body masses than their female counterparts. This is especially true for the Gentoo species. 
The average body mass data and table were generated using Python. R utilized Python handled data to create the plot.

